---
title: "IODS course project"
author: "Kalle Kivinen"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
---

***

# Introduction to Open Data Science - Week 1, Intro

```{r child = "chapter1.Rmd"}
# This will include Chapter 1 (that is updated in its own file) in the document.
```
  
***

# Introduction to Open Data Science - Week 2, Regression
  
```{r child = "chapter2.Rmd"}
# This will include Chapter 2 (that is updated in its own file) in the document.
```

***

# Introduction to Open Data Science - Week 3, Logistic Regression
  
```{r child = "chapter3.Rmd"}
# This will include Chapter 3 (that is updated in its own file) in the document.
```

***

# Introduction to Open Data Science - Week 4, Linear Discriminant Anlaysis and K-Means
  
```{r child = "chapter4.Rmd"}
# This will include Chapter 4 (that is updated in its own file) in the document.
```

***


# Introduction to Open Data Science - Week 5, Dimension Reduction Techniques
  
```{r child = "chapter5.Rmd"}
# This will include Chapter 5 (that is updated in its own file) in the document.
```

***

# Introduction to Open Data Science - Week 6, Analysis of Longitudinal Data
  
```{r child = "chapter6.Rmd"}
# This will include Chapter 5 (that is updated in its own file) in the document.
```

***